---
title: "Reproducible Research Project-1"
author: "Ammar Ahmed Alvi"
output:
  md_document:
    variant: markdown_github
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading required libraries

```{r libraries}

library(reshape2)
library(ggplot2)
library(plyr)

```


Loading and preparing the data

```{r Preparing}
data <- read.csv("activity.csv", header = TRUE)

data$date <- as.Date(data$date)


actMeltDate <- melt(data, id.vars="date", measure.vars="steps", na.rm=FALSE)
activity <- dcast(actMeltDate, date ~ variable, sum)

```

### Question 1: Calculating and plotting the mean and median of the total number of steps per day

You can also embed plots, for example:

```{r Q1a}

ggplot(activity, aes(x = date , y = steps)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline( yintercept = mean(activity$steps, na.rm = TRUE) , color = "Red", size = 1) +
  ggtitle("Daily Steps") +
  theme(plot.title = element_text(size = 18, hjust = 0.5))
  

```

Mean and Median

```{r Q1b}

paste("Mean steps per day = ", mean(activity$steps, na.rm = TRUE))  
paste("Median steps per day = ", median(activity$steps, na.rm = TRUE))

```


### Question 2: Calculating and plotting the average daily activity pattern by interval



```{r Q2a}

actMeltInt <- melt(data, id.vars="interval", measure.vars="steps", na.rm=TRUE)
actCastInt <- dcast(actMeltInt, interval ~ variable, mean)

ggplot(actCastInt, aes(x = interval , y = steps)) +
  geom_line(size= 1, color = "DarkRed") +
  geom_hline(yintercept = mean(actCastInt$steps), color = "Blue", size = 1) +
  ggtitle("Steps Taken at Each Interval") +
  theme(plot.title = element_text(size = 18, hjust = 0.5))


```




```{r Q2b}

paste("Interval with max value =", actCastInt$interval[which(actCastInt$steps == max(actCastInt$steps))])
paste("Maximum interval mean steps =", max(actCastInt$steps))

```


### Question 3: Imputing missing values to replace NAs in data set and compare results

```{r Q3a}

#Calculating the total number of missing values in the original data

sum(is.na(data$steps))

```

Since there are many missing values, I will replace them with the mean of their respective interval


```{r Q3b}

# Data frame with mean steps per interval - just renaming to be more descriptive
stepsInt <- actCastInt

# Create data frame that we will remove NAs from
CleanData <- data

# Merge activity data set with stepsPerInt data set
actMerge = merge(CleanData, stepsInt, by="interval", suffixes=c(".act", ".spi"))

# Get list of indexes where steps value = NA
naIndex = which(is.na(CleanData$steps))

# Replace NA values with value from steps.spi
CleanData[naIndex,"steps"] = actMerge[naIndex,"steps.spi"]

```



```{r Q3c}

actMeltDateNoNA <- melt(CleanData, id.vars="date", measure.vars="steps", na.rm=FALSE)
# Cast data frame to see steps per day
actCastDateNoNA <- dcast(actMeltDateNoNA, date ~ variable, sum)


ggplot(actCastDateNoNA, aes(x = date , y = steps)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_hline( yintercept = mean(actCastDateNoNA$steps, na.rm = TRUE) , color = "Red", size = 1) +
  ggtitle("Daily Steps (Imputted NAs)") +
  theme(plot.title = element_text(size = 18, hjust = 0.5))


```


Mean and Median

```{r Q3d}

paste("Mean steps per day = ", mean(actCastDateNoNA$steps, na.rm = TRUE))  
paste("Median steps per day = ", median(actCastDateNoNA$steps, na.rm = TRUE))

```



### Question 4: Check if differences exist in activity patterns between weekdays and weekends


Creating a factor variable to state whether a day is a weekend or not

```{r Q4a}

for (i in 1:nrow(CleanData)) {
    if (weekdays(CleanData$date[i]) == "Saturday" | weekdays(CleanData$date[i]) == "Sunday") {
        CleanData$dayOfWeek[i] = "weekend"
    } else {
        CleanData$dayOfWeek[i] = "weekday"
    }
}


```


Subsetting data to plot the graph

```{r Q4b}

# To create a plot, we must first subset the data
actWeekday <- subset(CleanData, dayOfWeek=="weekday")
actWeekend <- subset(CleanData, dayOfWeek=="weekend")

# Next, we need to process the data for our needs
actMeltWeekday <- melt(actWeekday, id.vars="interval", measure.vars="steps")
actMeltWeekend <- melt(actWeekend, id.vars="interval", measure.vars="steps")
actCastWeekday <- dcast(actMeltWeekday, interval ~ variable, mean)
actCastWeekend <- dcast(actMeltWeekend, interval ~ variable, mean)


```


Plotting graphs

```{r Q4c}

plot1 <- qplot(interval, steps, geom="line", data=actCastWeekday, type="bar", main="Steps by Interval - Weekday", xlab="Interval ID", ylab="Number of Steps")
plot2 <- qplot(interval, steps, geom="line", data=actCastWeekend, type="bar", main="Steps by Interval - Weekend", xlab="Interval ID", ylab="Number of Steps")

plot1
plot2

```







